### 2020-01-28 manuscript/main.tex
edits from August

August Guang$^1$, Mark Howison$^2$, Mia Coetzer$^3$, Lauren Ledingham$^3$, Matt D'Antuono$^3$,
Philip A. Chan$^3$, Charles Lawrence$^4$, Casey W. Dunn$^5$, Rami Kantor$^3$

$^1$ Computing and Information Services, Brown University, Providence, RI, USA

$^2$ Research Improving People's Lives, Providence, RI, USA

$^3$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA

@@

@@

While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing. In a meta study of published studies on HIV cluster infereance, 98 out of 105 (93\%) analyzed the \emph{pol} region \cite{hassan}.

@@

Most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny. Accordingly, researchers often summarize the within-host variation present in NGS data sets with a consensus sequence. In the larger context of phylogenetic methods (e.g. beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}. In the context of HIV, tihs assumption is that the consensus sequence adequately captures the relevant information about the distribution of sequences that might be constructed given the within-host variation measured in deep sequencing through NGS. Some previous studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.

In this study, we examine how well this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA. We present a new analysis method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.

@@

We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}. We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0). Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to all group M reference sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, and produces a codon frequency table across the near-whole HIV genome, which we refer to as the individual's HIV \emph{profile}.

We construct fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile. The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level. We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual. We perform phylogenetic inference on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}. In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences. We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance. Analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.

@@

@@

\begin{figure}[p!]
	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions. Numeric values indicate bootstrapped cluster support from the profile sampling method.}
	\centering
	\includegraphics[width=3.25in]{Figure6}
	\label{fig6}
\end{figure}

>>>>

August Guang$^{1,2}$, Mark Howison$^3$, Charles Lawrence$^4$, Casey W. Dunn$^5$, Rami Kantor$^6$

$^1$ Center for Computational Biology of Human Disease, Brown University, Providence, RI, USA

$^2$ Center for Computation and Visualization, Brown University, Providence, RI, USA

$^3$ Research Improving People's Lives, Providence, RI, USA

@@

$^6$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA

@@

While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing. In a meta study of published studies on HIV cluster inference, 98 out of 105 (93\%) analyzed the \emph{pol} region \parencite{hassan}.

@@

While the advantages of longer sequences in inferring transmission have been demonstrated, the advantages of deeper sequencing have not yet been fully investigated. In part this is because researchers often rely on consensus sequences, since most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny. Accordingly, researchers studying HIV transmission summarize the within-host HIV variation present in NGS data sets with a consensus sequence. In the larger context of phylogenetic methods (e.g. beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}. In the context of HIV, this assumption is that a consensus sequence can adequately capture all of the relevant information about within-host variation available in a deeply-sequenced NGS data set. Some previous studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.

In this study, we examine how well this assumption underlying the consensus approach holds on an NGS data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA. We present a new approach called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.

@@

We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}. We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0). A key feature of this pipeline is its use of profile hidden Markov models (HMMs) to model and align collections of HIV sequences. Profile HMMs have been used in many kinds of biological sequence analysis and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}. Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database with the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome. We refer to this resulting codon frequency table as the individual's HIV \emph{profile}.

We construct fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile. The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level. We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual. We construct the 500 full-resolved sequences in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the proflie representation of the individual's within-host variation.

We perform phylogenetic inference on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}. In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences with the same tools and parameters. We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance. Analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.

@@

\section*{Acknowledgments}

Mia Coetzer, Lauren Ledingham, Matt D'Antuono, Philip A. Chan

@@

\begin{figure}[p!]
	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions. Numeric values indicate bootstrapped cluster support from the profile sampling method. A blank cell indicates that the cluster was not detected in that consensus sequence and genomic region.}
	\centering
	\includegraphics[width=3.25in]{Figure6}
	\label{fig6}
\end{figure}
